\mytitle

### 2021-12-16 paper.tex
reduce bracket sizes
simplify equations
use interval [0,1/2]

As is well known, the performante of this naive implementation can be
improved by applying Horner's scheme, which reduces the number of
expensive floating\-/point multiplications to 7 and requires 6
additions:
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \Bigg\{1 + x \Bigg[-\frac{1}{2} + x \Bigg(\frac{1}{3} + x \Bigg\{-\frac{1}{4} \\
  &+ x \Bigg[\frac{1}{5} + x\Bigg(-\frac{1}{6} + \frac{x}{7}\Bigg)\Bigg]\Bigg\}\Bigg)\Bigg]\Bigg\}.
\end{split}\label{eq:horner}%
\end{align}
%
A C implementation using Horner's scheme could read
%
\lstinputlisting[language=C]{src/log1p_horner.c}
%
Note, that the formulation using Horner's scheme requires the
evaluation of the floating\-/point operations in strictly sequential
order.  However, the significant reduction of the number of
floating\-/point multiplications, compared to the naive
implementation, usually results in a shorter runtime.

There are several different possibilities to reduce the runtime of
Horner's method by rewriting the series expansion of $\ln(1+x)$ such
that the more terms can be calculated in parallel, while the number
floating\-/point operations is still kept small.  One possibility is
to split the r.h.s.\ of Eq.~\eqref{eq:horner} into two distinct sums,
one containing the even powers and the other containing the odd powers
of $x$,
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \Bigg\{1 + y \Bigg[\frac{1}{3} + y
  \Bigg(\frac{1}{5} + \frac{y}{7}\Bigg)\Bigg]\Bigg\} \\
  &+ y\Bigg[-\frac{1}{2} + y\Bigg(-\frac{1}{4} - \frac{y}{6}\Bigg)\Bigg],
\end{split}\label{eq:split}%
\end{align}
%
where $y=x^2$.  This formulation requires 8 floating\-/point
multiplications and 6 additions, i.e.\ one more multiplication than
Horner's scheme.  However, in Eq.~\eqref{eq:split} both sums can then
be calculated in parallel, leading to a potential speed-up if the cost
of the additional multiplication is smaller than the gain by the
parallel execution.  A C implementation could read:
%
\lstinputlisting[language=C]{src/log1p_split.c}
%
Another possibility is to rewrite the r.h.s.\ of Eq.~\eqref{eq:taylor}
using Estrin's scheme \cite{estrin},
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x + \left(\frac{x}{3}-\frac{1}{2}\right)
    y \\ &+\left[\left(\frac{x}{7}-\frac{1}{6}\right)
    y+\frac{x}{5}-\frac{1}{4}\right] z,
\end{split}\label{eq:estrin}%
\end{align}
%
where $y=x^2$ and $z=y^2$.  This form also requires 8 floating\-/point
multiplications and 6 additions, as in Eq.~\eqref{eq:split}.  However,
in Estrin's form more terms can be executed in parallel, leading to a
potential further speed-up.  A C implementation using Estrin's scheme
could read:
%
\lstinputlisting[language=C]{src/log1p_estrin.c}

@@

In \tabref{tab:runtime} the runtime of different C implementations is
compared on two different CPU architectures.  On each architecture the
implentations are compiled with \texttt{g++} with optimization
\texttt{-O2}.  The shown runtime denotes the total wall clock time for
the calculation of $\Li(x)$ for $10^6$ uniformly distributed random
values $x\in[-5,5]$.
%
\begin{table}[tb]
  \centering
  \caption{Runtime in seconds for the calculation of $\Li(x)$ for
    different implementations on different CPU architectures with
    different compilers.}
  \begin{tabular}{lcc}
    \toprule
    \multirow{2}{*}{Implementation} & i7-4700MQ & i7-5600U \\
    & \texttt{g++} 10.2.1 & \texttt{g++} 9.3.0 \\
    \midrule
    this paper & $0.031$ & $0.026$ \\
    Morris (DILOG 0011) \cite{morris} & $0.035$ & $0.026$ \\
    Cephes \cite{cephes} & $0.044$ & $0.034$ \\
    Algorithm 327 \cite{koelbigDilog} & $0.054$ & $0.035$ \\
    Chebyshev \cite{luke} & $0.079$ & $0.060$ \\
    Algorithm 490 \cite{ginsberg} & $0.125$ & $0.071$ \\
    GSL \cite{gsl} & $0.330$ & $0.155$ \\
    \bottomrule
  \end{tabular}
  \label{tab:runtime}
\end{table}
%
We find that the implementation presented in this paper and the
approach presented in \cite{morris} are fastest among the ones shown
in the table.  The next-to fastest implementation is provided by
Cephes, which is only $\sim 20\%$ slower.  The short runtime of Cephes
stems from the fact that Cephes also uses a rational function
approximation, where the numerator and the denominator can be
evaluated in parallel.  However, Cephes evaluates the polynomials in
the numerator and denominator using Horner's scheme, which is purely
sequential.  Furthermore, for $x>1$ Cephes calculates $\ln(x)$ twice,
which leads to a further slowdown.

>>>>

As is well known, the performante of this naive implementation can be
improved by applying Horner's scheme, which reduces the number of
expensive floating\-/point multiplications to 7 and requires 6
additions:
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \bigg\{1 + x \bigg[-\frac{1}{2} + x \bigg(\frac{1}{3} + x \bigg\{-\frac{1}{4} \\
  &+ x \bigg[\frac{1}{5} + x\bigg(-\frac{1}{6} + \frac{x}{7}\bigg)\bigg]\bigg\}\bigg)\bigg]\bigg\}.
\end{split}\label{eq:horner}%
\end{align}
%
A C implementation using Horner's scheme could read
%
\lstinputlisting[language=C]{src/log1p_horner.c}
%
Note, that the formulation using Horner's scheme requires the
evaluation of the floating\-/point operations in strictly sequential
order.  However, the significant reduction of the number of
floating\-/point multiplications, compared to the naive
implementation, usually results in a shorter runtime.

There are several different possibilities to reduce the runtime of
Horner's method by rewriting the series expansion of $\ln(1+x)$ such
that the more terms can be calculated in parallel, while the number
floating\-/point operations is still kept small.  One possibility is
to split the r.h.s.\ of Eq.~\eqref{eq:horner} into two distinct sums,
one containing the even powers and the other containing the odd powers
of $x$,
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \left\{1 + y \left[\frac{1}{3} + y
  \left(\frac{1}{5} + \frac{y}{7}\right)\right]\right\} \\
  &- y\left[\frac{1}{2} + y\left(\frac{1}{4} + \frac{y}{6}\right)\right],
\end{split}\label{eq:split}%
\end{align}
%
where $y=x^2$.  This formulation requires 8 floating\-/point
multiplications and 6 additions, i.e.\ one more multiplication than
Horner's scheme.  However, in Eq.~\eqref{eq:split} both sums can then
be calculated in parallel, leading to a potential speed-up if the cost
of the additional multiplication is smaller than the gain by the
parallel execution.  A C implementation could read:
%
\lstinputlisting[language=C]{src/log1p_split.c}
%
Another possibility is to rewrite the r.h.s.\ of Eq.~\eqref{eq:taylor}
using Estrin's scheme \cite{estrin},
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x + \left(\frac{x}{3}-\frac{1}{2}\right)
    y \\ &+\left[\left(\frac{x}{7}-\frac{1}{6}\right)
    y+\frac{x}{5}-\frac{1}{4}\right] z,
\end{split}\label{eq:estrin}%
\end{align}
%
where $y=x^2$ and $z=y^2$.  This form also requires 8 floating\-/point
multiplications and 6 additions, as in Eq.~\eqref{eq:split}.  However,
in Estrin's form more terms can be executed in parallel, leading to a
potential further speed-up.  A C implementation using Estrin's scheme
could read:
%
\lstinputlisting[language=C]{src/log1p_estrin.c}

@@

In \tabref{tab:runtime} the runtime of different C implementations is
compared on two different CPU architectures.  On each architecture the
implentations are compiled with \texttt{g++} with optimization
\texttt{-O2}.  The shown runtime denotes the total wall clock time for
the calculation of $\Li(x)$ for $10^6$ uniformly distributed random
values $x\in[0,1/2]$.
%
\begin{table}[tb]
  \centering
  \caption{Runtime in seconds for the calculation of $\Li(x)$ for
    different implementations on different CPU architectures with
    different compilers.}
  \begin{tabular}{lcc}
    \toprule
    \multirow{2}{*}{Implementation} & i7-5600U & ?? \\
    & \texttt{g++} 10.2.1 & \texttt{g++} 9.3.0 \\
    \midrule
    this paper                        & $0.0058$ & $??$ \\
    Morris (DILOG 0011) \cite{morris} & $0.0085$ & $??$ \\
    Cephes \cite{cephes}              & $0.0119$ & $??$ \\
    Algorithm 327 \cite{koelbigDilog} & $0.0172$ & $??$ \\
    Chebyshev \cite{luke}             & $0.0515$ & $??$ \\
    Algorithm 490 \cite{ginsberg}     & $0.0576$ & $??$ \\
    GSL \cite{gsl}                    & $0.0898$ & $??$ \\
    \bottomrule
  \end{tabular}
  \label{tab:runtime}
\end{table}
%
We find that the implementation presented in this paper and the
approach presented in \cite{morris} are fastest among the ones shown
in the table.  The next-to fastest implementation is provided by
Cephes, which is only $\sim 20\%$ slower.  The short runtime of Cephes
stems from the fact that Cephes also uses a rational function
approximation, where the numerator and the denominator can be
evaluated in parallel.  However, Cephes evaluates the polynomials in
the numerator and denominator using Horner's scheme, which is purely
sequential.  Furthermore, for $x>1$ Cephes calculates $\ln(x)$ twice,
which leads to a further slowdown.
